WFH and broadband speed (title needs rework)

### 2020-12-22 paper/v1_sage/broadband.speed.covid.tex
with amended appendix

Furthermore, these multi-layered digital divides intersect with material
divides and the economic geography of the UK. Following the regional
economic resilience literature, which underlines the differentiated
capacity of cities and regions to escape or recover from economic crises
\citep{martin2012regional, kitsos2018economic}, different places have
different industrial and occupational profiles, and these affect the
aggregated potential capacity of places for telecommuting. Such profiles
are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide
\citep{martin_north_south}. Various studies have illustrated severe
inequalities between the north and the south regions of England
\textbf{replace `UK'} regarding, for example, skills and human capital,
unemployment, productivity and prosperity
\citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
scholars have even argued that the UK suffers some of the highest level
of interregional inequalities in the global north
\citep{gal2018reducing, mccann2016uk}. Not only are all three levels of
digital divides associated, to a certain extent, with or shaped by the
geography of the UK, but the intersection of the digital and material
divides affects the capacity of places to overcome some of the economic
effects of the Covid-19 pandemic. Importantly, this is the first time
that digital technologies became an essential tool for economic
resilience for such a great part of the population.

@@

Meanwhile, those living in the largest cluster -- \(6\), with \(155\)
LADs home to \(29.5\) million people, experience aggregate mean upload
speeds of about \(1\)Mb/s faster than those in cluster \(13\), but still
lower than the other three large clusters and most of the smaller
clusters, suggesting a middling quality of service. However, the time
profile for cluster \(6\) in Figure \ref{UpClusterL} shows that upload
speeds drop quickly from \(6:00\) to \(7:00\) on a Monday and peak
between \(23:00\) and midnight on Wednesday and Thursday, but tend to be
lower during the working day. Indeed, experienced mean upload speeds in
the morning peak are \(4.4\)\% lower than in the evening peak -- a
greater, more noticeable change than any of the other large clusters
experience, but smaller than any of the clusters with temporal profiles
shown in Figure \ref{UpClusterS}. Clusters \(8\) and \(12\) also have
mean upload speeds under \(10\)Mb/s, but higher than clusters \(1\) and
\(13\), and their standard deviation is not dissimilar. However, this
masks great variation in when lower speeds are experienced, with the
mean upload speeds much lower between \(9:00-10:59\) than between
\(19:00-20:59\) in cluster \(8\), but slightly faster in the morning in
cluster \(12\).

@@

Indeed cluster \(12\) is the only cluster to experience higher speeds in
the evening peak, compared to the morning peak, suggesting that
widespread telecommuting has generally changed the temporal profile of
internet activity throughout the UK. Yet even if all but one cluster is
showing slower speeds in the morning than the evening, the reliability
of internet services in different clusters during the working day still
varies considerably. Interpreting this variation from the large spikes
and dips shown on Figures \ref{UpClusterL} and \ref{UpClusterS} is
difficult, but the statistics in Table \ref{up.cluster.descr} show that
clusters \(9\) and \(11\) have the most reliable internet services. The
ratio of standard deviation to mean in both these clusters is below
55\%, and the ratio of upload speeds in the peak periods indicates that
speeds are only about 2\% slower in the morning. Mean speeds are also
higher than in any other cluster, excluding cluster \(2\), where
measures of reliability suggests poorer performance.

@@

In summary, the local authorities in clusters \(9\) and \(11\)
experienced resilient broadband internet that could support high levels
of telecommuting. Those in clusters \(2\), \(3\), \(7\), and \(10\) also
experience higher than average mean speeds and rank high to middle on
measures of service reliability. These LADs are \emph{not} on the wrong
side of the first level digital divide, but how likely are they to be
able to take advantage of their resilient ICT infrastructure and
services? Meanwhile, cluster \(6\) is not only the largest in terms of
number of LADs and population, it has the closest mean upload speed to
the pre-clustered average for the whole sample. As well as average
quality internet services, those in cluster \(6\) also experience
average reliability for work purposes, ranking fifth behind the four
other clusters with populations over one million, but ahead of the
smaller clusters. Clusters \(8\) and \(12\) are also close to average
mean upload speeds, but show very different patterns in terms of
reliability, whilst clusters \(1\) and \(13\) appear to suffer most from
a lack of quality internet services, with slow speeds and high standard
deviations. With those in cluster \(1\) in particular more likely to
experience that poor reliability during the morning peak, is this first
level digital divide occurring in areas where few are occupationally
able to telecommute anyway, and what are the implications for economic
resilience?

@@

Using auxiliary regressions, we test whether the clusters that have
higher mean speeds and more reliable services consist of LADs that are
more urban, affluent, and / or more likely to benefit from a choice of
high quality internet services. We also estimate which of our clusters
are more likely to have a higher proportion of occupations which enable
telecommuting because of the nature of the work. The results of these
auxiliary regressions are presented in Table \ref{aux}. The dependent
variable is the LAD cluster membership as described in the
\protect\hyperlink{sec:3}{methods and data} section and equation
\ref{eq1} and each column represent a different cluster. The reference
case is cluster \(4\), which includes only the local authority of
Hambleton in North Yorkshire, a rural area of just over ninety thousand
people. Mean, experienced upload speeds in cluster \(4\) are close to
both the average speeds for the \(13\) clusters (\(9.9\)Mb/s) and the
pre-clustered average for the whole sample (\(9.3\)Mb/s) as shown in
\ref{up.cluster.descr}. However, the standard deviation for cluster
\(4\) and the difference between average speeds in the morning compared
to the evening peak periods are indications of worse reliability than
many of the other clusters. Hence, the results in Table \ref{aux} should
be seen as relative rather than absolute probabilities.

First, the number of speed tests run per cluster inhabitant between
\(9:00\)-\(10:59\) is an indication of satisfaction or at least a lack
of concern over broadband speeds and quality of service. People in LADs
assigned to clusters \(6\) and \(13\) were more likely to ran speed
tests (at a per capita basis) at this time of day, whilst people in
cluster \(11\) were the least likely ones. This high level of
satisfaction matches the high quality and reliability of service we
would expect from the analysis in Section {[}4.1{]}{]}(\#sec:4.1). Those
in clusters \(2\), \(3\), \(9\) and \(10\) were also less likely to ran
speed tests per capita than those in the other six multi-authority
clusters. Cluster \(7\) was the only cluster with higher mean speeds and
middle to high measures of reliability which was more likely to run more
speed tests per capita. This might be a reflection of the choice of ISP
and connection options in those LADs, as, along with cluster \(12\),
those in cluster \(7\) were least likely to have Virgin Media
connections. Fast Virgin Media cable connections have historically only
been available to \(45\)\% of premises in the UK \citep{ofcom2016},
where the more lucrative and competitive market originally attracted the
cable TV provider. Those in clusters \(2\), \(3\), \(9\) and \(11\)
benefit from a higher proportion of Virgin connections, which is an
indication that people in these clusters are more likely to live in
urban areas, with more choice of broadband services, or in other words,
are more likely to be on the right side of the first level,
infrastructure-based digital divide.

We employ distance from London and for the nearest metropolitan area
(including London) as two more variables depicting peripherality, urban
structure and, potentially, first level digital divides. The broadband
speed tests run in the authorities in cluster \(3\) are more likely to
be taking place close to London than those run in any of the other
clusters, a result that makes sense considering two of the four
authorities in cluster \(3\) are the London commuter towns of Harlow and
Luton. However, even though London was also included in the variable
calculating distance from the centre of one of either the ten largest
metropolitan areas in England, Glasgow or Cardiff, tests run in cluster
\(3\) are likely to be furthest away. Still, whilst significant, the
effect is tiny and the other members of cluster \(3\) are Corby and
Eastbourne, two towns which, whilst not part of metropolitan areas, are
also not in remote parts of the country, and are home to an estimated
population of \(175\) thousand people between them. \textbf{move maps to
here?} Meanwhile, LADs in cluster \(7\) are most likely to be near the
centre of a large metropolitan area, even though the four local
authorities of cluster \(7\) include no central urban boroughs, and only
one LAD that is part of a metropolitan area of governance - Tameside in
Greater Manchester. This may be why those in cluster \(7\) are not
served by Virgin Media. It also is a demonstration of the complexity of
experienced broadband upload speeds as captured by time-series
clustering, and their likely interaction with first level digital
divides. Cluster \(1\), which our analysis suggests lacks broadband
resilience, contains five, mainly rural authorities. However, they are
scattered around the country -- Ceredigion in West Wales, Darlington in
the Northeast, Eden between the Lake District and the North Penines,
Rossendale in the Northwest, and Rutland in the East Midlands -- and
therefore, the results in Table \ref{aux} indicate that these
authorities are not as far from the centre of a metropolitan area as
authorities in cluster \(3\), but furthest from London.

Internet resilience, as measured here, seems to be more nuanced than our
-- arguably crude -- dummy variable depicting the North-South economic
divide either, which assigns \(1\) to LADs locates in Greater London,
Southeast, Southwest and East of England regions. The authorities most
likely to be in the South are those in cluster \(13\), which was
identified as having the slowest mean upload speeds of any of the
clusters, and thus a low level of service. However, cluster \(13\) does
include some rural, remote areas of the country, such as Northwest
Scotland, Cornwall and Powys in Wales as shown on Figure
\ref{map.up.clusters}. It also includes the major metropolitan centres
of Bristol, Liverpool, and Newcastle, as well as nine (of \(32\)) London
Boroughs and plenty of home county and suburban areas. Although by the
standard deviation measure cluster \(13\) has unreliable internet
services, speed variation was less during the morning peak. More
broadly, the estimates for this variable are not consistent between the
two clubs of LAD clusters: those who are in the wrong and the right side
of digital divides.

Regarding the rest of the socio-economic variables we see a varied
picture. Clusters \(9\) and \(11\), which are the ones benefiting by the
most resilient internet connections are likely to consist of LADs with
higher share of managerial and highly skilled occupations as well as
higher average earnings. Interestingly, both clusters are characterised
by lower share of occupations that are not usually associated with
teleworking (professional, associated professional and technical,
administrative, leisure and machine operation jobs), but also by lower
number of business per habitant and. However, the likelihood of hosting
highly skilled residents is not consistent with cluster \(11\) having a
negative coefficient for the NVQ4 variable. This might be an indication
of the higher returns that resilient broadband infrastructure generates
for cluster \(9\) versus cluster \(11\). While both of these cluster
consist more likely by large LADs in terms of population, job density if
higher for LADs in cluster \(11\) indicating more labour intensive
economic structures contrary to LADs in cluster \(9\).

A similar divided picture can be built for the next tier of clusters:
\(2\), \(3\), \(7\), \(10\). LADs in these clusters experience higher
than average mean speeds and rank high to middle on measures of service
reliability. As mentioned before, LADs in these clusters are on the
right side of the first level digital divide. However, while all of
them are more likely to be populated by people employed in managerial
jobs, only LADs in cluster \(2\) and \(10\) are more likely to have
highly skilled work force, which can more easily take advantage of
working from home opportunities. A more mixed picture can be built for
LADs in clusters \(3\) and \(7\), which are associated, on the one hand,
with lower skills and, on the other hand, with higher likelihood of
individuals in managerial occupations. A similar contradicting pattern
can be illustrated for cluster \(6\), which is the largest one with
average quality of internet service. Arguably, LADs in clusters \(2\)
and \(10\) are the ones benefiting more by resilient internet
infrastructure in an era that working from home became a vehicle for
economic resilience.

Moving on to the next tier of clusters \(8\) and \(12\), which are
characterised by LADs enjoying close to average mean upload speeds, but
different patterns of internet reliability, we see a differentiated
picture. On the one hand, both of them are likely to host individuals
employed in managerial occupations while the opposite applies on
skilled, professional and leisure occupations. On the other hand, LADs
in cluster \(8\) are characterised by a very small positive and LADs in
cluster \(12\) by a negative likelihood to host highly skilled
individuals. Moreover, LADs in these two clusters have a rather similar
profile regarding the occupational variables, but different tendencies
regarding the stock of businesses. In total, LADs in these clusters do
not benefit by great internet services and could only marginally benefit
by working from home opportunities.

Lastly, clusters \(1\) and \(13\), which appear to suffer most from a
lack of quality internet services are more likely to also suffer from
low skilled workforce. LADs in cluster \(13\). Contradictory, the
likelihood of workforce employed in managerial occupations is positive
for LADs in both clusters, but only individuals in LADs in cluster
\(13\) are more likely to enjoy high earnings.

@@

those in cluster \(6\), including people living in Birmingham, Leeds and
Sheffield, in twelve London Boroughs, and many suburban areas and
smaller cities like Cardiff, Oxford and Cambridge This lower level of
reliability may be as a result of increased demand, e.g.~for
telecommuting, especially as there are few truly remote areas in this
cluster.

The eight LADs in clusters \(3\) and \(10\) are home to over one million
people, and are all more urban and suburban than cluster \(1\),
including the London Borough of Newham, the Southeast commuter towns of
Harlow and Luton, the Birmingham suburbs of Bromsgrove and Cannock
Chase, Dundee City, and the towns of Eastbourne and Corby.

Although not all of the local authorities in these clusters \(9\) and
\(11\) are urban, among the \(65\) LADs are Manchester and three of the
nine other boroughs of Greater Manchester, Glasgow, Nottingham, both
Portsmouth and Southampton, eight London Boroughs, four of the seven
constituent authorities of the West Midlands Combined Authority. There
are also a number of other tightly bounded urban areas, such as
Aberdeen, Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban
areas at the centre of less confined districts like Burnley, Milton
Keynes, and Northampton.

@@

\hypertarget{sec:appendix}{%
\section{Appendix}\label{sec:appendix}}

\begin{table}[!htbp] \centering 
  \caption{LAD cluster membership for upload speed timeseries\label{cluster.membership}} 
\footnotesize 
\begin{tabular}{@{\extracolsep{5pt}} ccc} 
 & LAD & cluster membership \\ 
1 & Aberdeen City & 9 \\ 
2 & Aberdeenshire & 13 \\ 
3 & Adur & 13 \\ 
4 & Allerdale & 6 \\ 
5 & Amber Valley & 6 \\ 
6 & Angus & 6 \\ 
7 & Antrim and Newtownabbey & 13 \\ 
8 & Ards and North Down & 12 \\ 
9 & Argyll and Bute & 13 \\ 
10 & Armagh City, Banbridge and Craigavon & 13 \\ 
11 & Arun & 13 \\ 
12 & Ashfield & 6 \\ 
13 & Ashford & 6 \\ 
14 & Aylesbury Vale & 6 \\ 
15 & Babergh & 13 \\ 
16 & Barking and Dagenham & 11 \\ 
17 & Barnet & 6 \\ 
18 & Barnsley & 9 \\ 
19 & Barrow-in-Furness & 13 \\ 
20 & Basildon & 13 \\ 
21 & Basingstoke and Deane & 6 \\ 
22 & Bassetlaw & 13 \\ 
23 & Bath and North East Somerset & 6 \\ 
24 & Bedford & 13 \\ 
25 & Belfast & 6 \\ 
26 & Bexley & 6 \\ 
27 & Birmingham & 6 \\ 
28 & Blaby & 11 \\ 
29 & Blackburn with Darwen & 13 \\ 
30 & Blackpool & 11 \\ 
31 & Blaenau Gwent & 6 \\ 
32 & Bolsover & 11 \\ 
33 & Bolton & 13 \\ 
34 & Boston & 13 \\ 
35 & Bournemouth, Christchurch and Poole & 6 \\ 
36 & Bracknell Forest & 13 \\ 
37 & Bradford & 6 \\ 
38 & Braintree & 6 \\ 
39 & Breckland & 13 \\ 
40 & Brent & 11 \\ 
41 & Brentwood & 6 \\ 
42 & Bridgend & 6 \\ 
43 & Brighton and Hove & 13 \\ 
44 & Bristol, City of & 13 \\ 
45 & Broadland & 13 \\ 
46 & Bromley & 6 \\ 
47 & Bromsgrove & 10 \\ 
48 & Broxbourne & 9 \\ 
49 & Broxtowe & 6 \\ 
50 & Burnley & 11 \\ 
51 & Bury & 6 \\ 
52 & Caerphilly & 11 \\ 
53 & Calderdale & 6 \\ 
54 & Cambridge & 6 \\ 
55 & Camden & 13 \\ 
56 & Cannock Chase & 10 \\ 
57 & Canterbury & 6 \\ 
58 & Cardiff & 6 \\ 
59 & Carlisle & 11 \\ 
60 & Carmarthenshire & 13 \\ 
61 & Castle Point & 6 \\ 
62 & Causeway Coast and Glens & 11 \\ 
63 & Central Bedfordshire & 13 \\ 

>>>>

Furthermore, these multi-layered digital divides intersect with material
divides and the economic geography of the UK. Following the regional
economic resilience literature, which underlines the differentiated
capacity of cities and regions to escape or recover from economic crises
\citep{martin2012regional, kitsos2018economic}, different places have
different industrial and occupational profiles, and these affect the
aggregated potential capacity of places for telecommuting. Such profiles
are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide
\citep{martin_north_south}. Various studies have illustrated severe
inequalities between the north and the south regions of England
regarding, for example, skills and human capital, unemployment,
productivity and prosperity
\citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
scholars have even argued that the UK suffers some of the highest level
of interregional inequalities in the global north
\citep{gal2018reducing, mccann2016uk}. Not only are all three levels of
digital divides associated, to a certain extent, with or shaped by the
geography of the UK, but the intersection of the digital and material
divides affects the capacity of places to overcome some of the economic
effects of the Covid-19 pandemic. Importantly, this is the first time
that digital technologies became an essential tool for economic
resilience for such a great part of the population.

@@

Meanwhile, those living in the largest cluster -- \(6\), with \(155\)
LADs home to \(29.5\) million people, experience aggregate mean upload
speeds of about \(1\)Mb/s faster than those in cluster \(13\), but still
lower than the other three large clusters and most of the smaller
clusters, suggesting a middling quality of service. However, the time
profile for cluster \(6\) in Figure \ref{UpClusterL} shows that upload
speeds are highest at \(6:00\) on a Monday morning and peak between
\(23:00\) and midnight on Wednesday and Thursday, but tend to be lower
during the working day. Indeed, experienced mean upload speeds in the
morning peak are \(4.4\)\% lower than in the evening peak -- a greater,
more noticeable change than any of the other large clusters experience,
but smaller than any of the clusters with temporal profiles shown in
Figure \ref{UpClusterS}. Clusters \(8\) and \(12\) also have mean upload
speeds under \(10\)Mb/s, but higher than clusters \(1\) and \(13\), and
their standard deviation is not dissimilar. However, this masks great
variation in when lower speeds are experienced, with the mean upload
speeds much lower between \(9:00-10:59\) than between \(19:00-20:59\) in
cluster \(8\), but slightly faster in the morning in cluster \(12\).

@@

Indeed cluster \(12\) is the only cluster to experience higher speeds in
the evening peak, compared to the morning peak, suggesting that
widespread telecommuting has generally changed the temporal profile of
internet activity throughout the UK. Yet even if all but one cluster is
showing slower speeds in the morning than the evening, the reliability
of internet services in different clusters during the working day still
varies considerably. Interpreting this variation from the large spikes
and dips shown on Figures \ref{UpClusterL} and \ref{UpClusterS} is
difficult, but the statistics in Table \ref{up.cluster.descr} show that
clusters \(9\) and \(11\) have the most reliable internet services. The
ratio of standard deviation to mean in both these clusters is below
55\%, and the ratio of upload speeds in the peak periods indicates that
speeds are only about 2\% slower in the morning than the evening. Mean
speeds are also higher than in any other cluster, excluding cluster
\(2\), where measures of reliability suggests poorer performance.

@@

In summary, the local authorities in clusters \(9\) and \(11\)
experienced resilient broadband internet that could support high levels
of telecommuting. Those in clusters \(2\), \(3\), \(7\), and \(10\) also
experience higher than average mean speeds and rank middle to high on
measures of service reliability. These LADs are \emph{not} on the wrong
side of the first level digital divide, but how likely are they to be
able to take advantage of their resilient ICT infrastructure and
services? Meanwhile, cluster \(6\) is not only the largest in terms of
number of LADs and population, it has the closest mean upload speed to
the pre-clustered average for the whole sample. As well as average
quality internet services, those in cluster \(6\) also experience
average reliability for work purposes, ranking fifth behind the four
other clusters with populations over one million, but ahead of the
smaller clusters. Clusters \(8\) and \(12\) are also close to average
mean upload speeds, but show very different patterns in terms of
reliability, whilst clusters \(1\) and \(13\) appear to suffer most from
a lack of quality internet services, with slow speeds and high standard
deviations. With those in cluster \(1\) in particular more likely to
experience that poor reliability during the morning peak, is this first
level digital divide occurring in areas where few are occupationally
able to telecommute anyway, and what are the implications for economic
resilience?

@@

Using auxiliary regressions, we test whether the clusters that have
higher mean speeds and more reliable services consist of LADs that are
more urban, affluent, and / or more likely to benefit from a choice of
high quality internet services. We also estimate which of our clusters
are more likely to have a higher proportion of occupations which enable
telecommuting because of the nature of the work. The results of these
auxiliary regressions are presented in Table \ref{aux}. The dependent
variable is the LAD cluster membership as described in the
\protect\hyperlink{sec:3}{methods and data} section and equation
\ref{eq1} and each column represent a different cluster. The reference
case is cluster \(4\), which includes only the local authority of
Hambleton in North Yorkshire, a rural area of just over ninety thousand
people. Mean, experienced upload speeds in cluster \(4\) are close to
both the average speeds for the \(13\) clusters (\(9.9\)Mb/s) and the
pre-clustered average for the whole sample (\(9.3\)Mb/s) as shown in
Table \ref{up.cluster.descr}. However, the standard deviation for
cluster \(4\) and the difference between average speeds in the morning
compared to the evening peak periods are indications of worse
reliability than many of the other clusters. Hence, the results in Table
\ref{aux} should be seen as relative rather than absolute probabilities.

First, the number of speed tests run per cluster inhabitant between
\(9:00\)-\(10:59\) is an indication of satisfaction or at least a lack
of concern over broadband speeds and quality of service. People in LADs
assigned to clusters \(6\) and \(13\) were more likely to run speed
tests (at a per capita basis) at this time of day, whilst people in
cluster \(11\) were the least likely to do so. This high level of
satisfaction matches the high quality and reliability of service we
would expect from the analysis in Section {[}4.1{]}{]}(\#sec:4.1). Those
in clusters \(2\), \(3\), \(9\) and \(10\) were also less likely to ran
speed tests per capita than those in the other six multi-authority
clusters. Cluster \(7\) was the only cluster with higher mean speeds and
middle to high measures of reliability which was more likely to run more
speed tests per capita. This might be a reflection of the choice of ISP
and connection options in those LADs, as, along with cluster \(12\),
those in cluster \(7\) were least likely to have Virgin Media
connections. Fast Virgin Media cable connections have historically only
been available to \(45\)\% of premises in the UK \citep{ofcom2016},
where the more lucrative and competitive market originally attracted the
cable TV provider. Those in clusters \(2\), \(3\), \(9\) and \(11\)
benefit from a higher proportion of Virgin connections, which is an
indication that people in these clusters are more likely to live in
urban areas, with more choice of broadband services, or in other words,
are more likely to be on the right side of the first level,
infrastructure-based digital divide.

We employ distance from London and for the nearest metropolitan area
(including London) as two more variables depicting peripherality, urban
structure and, potentially, first level digital divides. The broadband
speed tests run in the authorities in cluster \(3\) are more likely to
be taking place close to London than those run in any of the other
clusters, a result that makes sense considering two of the four
authorities in cluster \(3\) are the London commuter towns of Harlow and
Luton. However, even though London was also included in the variable
calculating distance from the centre of one of either the ten largest
metropolitan areas in England, or Glasgow or Cardiff, tests run in
cluster \(3\) are likely to be furthest away. Still, whilst significant,
the effect is tiny and the other members of cluster \(3\) are Corby and
Eastbourne, two towns which, whilst not part of metropolitan areas, are
also not in remote parts of the country, and are home to an estimated
population of \(175\) thousand people between them. \textbf{move maps to
here?}

Meanwhile, LADs in cluster \(7\) are most likely to be near the centre
of a large metropolitan area, even though the four local authorities of
cluster \(7\) include no central urban boroughs and only one LAD that is
part of a metropolitan area of governance - Tameside in Greater
Manchester. This may be why those in cluster \(7\) are not served by
Virgin Media. It also is a demonstration of the complexity of
experienced broadband upload speeds as captured by time-series
clustering, and their likely interaction with first level digital
divides. Cluster \(1\), which our analysis suggests lacks broadband
resilience, contains five, mainly rural authorities. However, they are
scattered around the country -- Ceredigion in West Wales, Darlington in
the Northeast, Eden between the Lake District and the North Penines,
Rossendale in the Northwest, and Rutland in the East Midlands -- and
therefore, the results in Table \ref{aux} indicate that these
authorities are not as far from the centre of a metropolitan area as
authorities in cluster \(3\), but furthest from London.

Internet resilience, as measured here, seems to be more nuanced than our
-- arguably crude -- dummy variable depicting the North-South economic
divide either, which assigns \(1\) to LADs locates in Greater London,
Southeast, Southwest and East of England regions, and \(0\) to the rest.
The authorities most likely to be in the South are those in cluster
\(13\), which was identified as having the slowest mean upload speeds of
any of the clusters, and thus a low level of service. However, cluster
\(13\) does include some rural, remote areas of the country, such as
Northwest Scotland, Cornwall and Powys in Wales as shown on Figure
\ref{map.up.clusters}. It also includes the major metropolitan centres
of Bristol, Liverpool, and Newcastle, as well as nine (of \(32\)) London
Boroughs, including the City of London and plenty of home county and
suburban areas. Although by the standard deviation measure cluster
\(13\) has unreliable internet services, speed variation was less during
the morning peak, suggesting that the estimates for reliability are
inconsistent. Considering that this is one of the largest clusters, and
thus the averages incorporate more noise than some of the smaller
clusters, it may be that the LADs in this cluster do not all suffer
equally from a first level digital divide.

Yet we need to consider the results for other variables in Table
\ref{aux} to better determine whether clusters \(1\) and \(13\), which
appear to suffer most from a lack of quality internet services, are also
more likely to have a low skilled workforce, less able to benefit from
telecommuting. Cluster \(1\) has the lowest propotion of educated people
of any of the clusters, and the lowest earnings, despite recording the
highest proportion of managerial, professional jobs, and the second
highest proportion of tech jobs. Rural areas such as those in cluster
\(1\) are often home to many retired people, which might explain these
results or perhaps, as these figures are relative, we could note that
cluster \(1\) also has a greater proportion of skilled trades than other
clusters. Either way, it seems likely that the first level digital
divide reinforces other inequities in cluster \(1\). In comparison,
those cluster \(13\) are more likely to earn more and have a better
education. Cluster \(13\) also has the most businesses per inhabitant,
but is somewhere in the middle in terms of job density. If this is an
indication of a high number of SMEs, it might give some explanation for
the variable internet quality, considering that small businesses have
not been seen as the most valuable customers for higher speed broadband
services \textbf{insert ref}.

Cluster \(6\) is the largest cluster, and thus, like cluster \(13\),
there is more noise within the averages we use to measure internet
quality and reliability, and the result indicates an average level of
service, even if better than LADs in cluster \(13\). Cluster \(6\) also
falls towards the middle of the clusters on many of the socio- economic
variables in Table \ref{aux}. It ranks third or fourth out of the eleven
clusters in the table in terms of having a higher proportion of
managerial, professional, and tech jobs, as well as higher earnings, but
is fourth from bottom in terms of education. The LADs in cluster \(6\)
are also diverse, with few truly remote areas, but many urban areas in
both the North, including Birmingham, Leeds and Sheffield, and South,
including twelve London Boroughs, and many suburban areas and smaller
cities like Oxford and Cambridge. The capital cities of the other UK
nations, Belfast, Cardiff and Edinburgh, are also in this cluster,
suggesting perhaps that the lower level of reliability discussed in
Section \protect\hyperlink{sec:4.1}{4.1} may be as a result of increased
demand, e.g.~for telecommuting.

In comparison, clusters \(9\) and \(11\), which benefit from the most
resilient internet connections, are characterised by lower share of
occupations that are\\
may benefit from the ability to telecommute (e.g.~managerial and
technical). Cluster \(9\) is more highly educated and higher earning
than cluster \(11\), has fewer businesses per inhabitant and less job
density. Indeed, cluster \(11\) has a negative coefficient for the NVQ4
variable, whilst for cluster \(9\) the coefficient is positive, and
cluster \(11\) has the second highest job density and cluster \(9\) the
second lowest. The former might be an indication of the higher returns
that resilient broadband infrastructure generates for cluster \(9\)
versus cluster \(11\). Indeed, we did detect slightly more slowdown in
the AM peak in cluster \(9\) than in cluster \(11\) (see Table
\ref{up.cluster.descr}). Meanwhile, a propensity for a high job density
might indicate more labour intensive economic structures in cluster
\(11\) than in the LADs in cluster \(9\).

The LADs in clusters \(9\) and \(11\) also have opposite signs for their
likelihood of being in the South of England. Those in cluster \(9\) less
likely to be in the South and those in cluster \(11\) more likely,
although the coefficient for cluster \(9\) might be skewed by the
presence of the Scottish cities of Glasgow and Aberdeen, as Scotland has
a different economic profile than England. Still, both clusters consist
of larger LADs in terms of population, including the urban centre of
Manchester and three of the nine other districts of Greater Manchester,
Nottingham, both Portsmouth and Southampton, eight London Boroughs, and
four of the seven constituent authorities of the West Midlands Combined
Authority. Among the \(65\) LADs are also a number of other tightly
bounded urban areas, such as Blackpool, Ipswich, Norwich, Slough, and
Stevenage, and urban areas at the centre of less confined districts like
Burnley, the digital hub of Milton Keynes, and Northampton. Thus, our
findings confirm the importance of geography and urban locations for
being on the right side of the first level digital divide, and suggest
that internet resilience cuts across other digital and economic divides.

A similarly divided picture can be built for the next tier of clusters:
\(2\), \(3\), \(7\), \(10\). LADs in these clusters experience higher
than average mean speeds and rank high to middle on measures of service
reliability. LADs in these clusters are on the right side of the first
level digital divide. However, only LADs in cluster \(2\) and \(10\) are
more likely to have highly skilled work force, which might more easily
take advantage of working from home opportunities. Cluster \(2\) is
comprised of just two LADs, with the lowest job density of any cluster
-- these are East Lothian near Edinburgh, and North East Lincolnshire,
home to Grimsby. Cluster \(10\) is comprised of four LADs, including two
peripheral suburban areas of Birmingham, the East London Borough of
Newham, and Dundee. Suburbs are considered the most likely urban form in
which telecommuters live, and Dundee has a reputation for tech startups
\textbf{insert ref}. Yet although those in clusters \(2\) and \(10\) may
be making the most of their resilient internet, those in cluster \(7\)
have the highest earnings. The four LADs in cluster \(7\) include the
suburbs South of Belfast, a suburban district of Greater Manchester, the
villages between York and Leeds, and a couple of small West Midlands
towns. LADs in cluster \(3\) are associated more with lower skills and
have the highest proportion of individuals in machine operation jobs.
They include the Southeast commuter towns of Harlow and Luton and the
slightly more distant towns of Eastbourne and Corby. Arguably, LADs in
clusters \(3\) benefit the least from their resilient internet
infrastructure in an era when working from home became a vehicle for
economic resilience.

Finally, clusters \(8\) and \(12\), are characterised by LADs enjoying
close to average mean upload speeds, but different patterns of internet
reliability. LADs in cluster \(8\) are likely to host more individuals
employed in tech and administrative occupations than any other cluster,
as well as many in managerial occupations, whilst the opposite applies
to skilled trade and leisure occupations. These LADs are characterised
by a very small positive likelihood of more educated residents, but are
not significantly likely to earn more than other clusters. Including
suburbs of Leicester, Cardiff, Birmingham, the green belt area south of
London straddling the M25 motorway, and rural West Devon, the LADs in
cluster \(8\) seem to likely to have the skills and occupations that
would benefit from quality internet services, but suffer from poor
internet resilience and reliability.

In comparison, LADs in cluster \(12\) are less likely to be able to
benefit from quality internet services if they had them, with fewer
individuals achieving NVQ4 or better and lower levels of occupations
that would benefit from homeworking. Yet LADs in this cluster benefit
from the second highest level of earnings, and have the largest stock of
businesses per inhabitant and the highest job density. This density of
businesses could be why cluster \(12\) is the only cluster with higher
speeds in the morning peak during the study period. If people are at
home, business premises might well be abandoned. Cluster \(12\) is home
to \(1.5\)million people spread across twelve LADs, including the City
of Westminster and the London Borough of Hammersmith. However, it also
includes a LAD in Northern Ireland, two in Wales, two suburban areas
North of London, Preston in the Northwest, East Staffordshire in the
Midlands, and Fenland in the East of England. This spatial diversity
demonstrates that the temporal clustering of internet resilience does
not create spatial clusters, and digital divides do not necessarily
overlap with economic ones.

@@

@@

\pagebreak

\hypertarget{appendix1}{%
\section{Appendix 1}\label{appendix1}}

This is the LAD cluster membership for the upload speed timeseries.

\textbf{Cluster 1: } Ceredigion, Darlington, Eden, Rossendale, Rutland

\textbf{Cluster 2: } East Lothian, North East Lincolnshire

\textbf{Cluster 3: } Corby, Eastbourne, Harlow, Luton

\textbf{Cluster 4: } Hambleton

\textbf{Cluster 5: } Fylde

\textbf{Cluster 6: } Allerdale, Amber Valley, Angus, Ashfield, Ashford,
Aylesbury Vale, Barnet, Basingstoke and Deane, Bath and North East
Somerset, Belfast, Bexley, Birmingham, Blaenau Gwent, Bournemouth,
Christchurch and Poole, Bradford, Braintree, Brentwood, Bridgend,
Bromley, Broxtowe, Bury, Calderdale, Cambridge, Canterbury, Cardiff,
Castle Point, Chelmsford, Cheltenham, Cherwell, Chesterfield, City of
Edinburgh, Clackmannanshire, Colchester, Copeland, County Durham,
Coventry, Croydon, Dartford, Daventry, Denbighshire, Derby, Derry City
and Strabane, Dorset, Ealing, East Ayrshire, East Hampshire, East
Lindsey, East Northamptonshire, East Renfrewshire, East Riding of
Yorkshire, East Suffolk, Eastleigh, Elmbridge, Enfield, Falkirk,
Fareham, Gateshead, Gedling, Gosport, Gravesham, Great Yarmouth,
Guildford, Harborough, Haringey, Harrogate, Harrow, Hart, Hartlepool,
Havering, Herefordshire, County of, High Peak, Hinckley and Bosworth,
Horsham, Islington, Kettering, King's Lynn and West Norfolk, Kingston
upon Thames, Kirklees, Leeds, Leicester, Lincoln, Maidstone, Maldon,
Mansfield, Medway, Mendip, Mid Sussex, Middlesbrough, Monmouthshire,
Moray, New Forest, Newcastle-under-Lyme, Newport, North Ayrshire, North
East Derbyshire, North Hertfordshire, North Kesteven, North Lanarkshire,
North Lincolnshire, North Norfolk, North Tyneside, North West
Leicestershire, Northumberland, Nuneaton and Bedworth, Oxford,
Pembrokeshire, Pendle, Renfrewshire, Ribble Valley, Rochford, Runnymede,
Rushcliffe, Ryedale, Salford, Sefton, Sheffield, Shropshire, Solihull,
South Ayrshire, South Hams, South Holland, South Lanarkshire, South
Oxfordshire, South Staffordshire, St Albans, Staffordshire Moorlands,
Stockport, Stoke-on-Trent, Surrey Heath, Sutton, Swale, Tamworth,
Tendring, Test Valley, Thurrock, Tonbridge and Malling, Torfaen,
Wakefield, Warrington, Warwick, Wealden, Wellingborough, West Berkshire,
West Dunbartonshire, West Lancashire, West Lothian, West Oxfordshire,
West Suffolk, Wigan, Wiltshire, Woking, Worcester, Wrexham, Wycombe,
York
